Accelerating Taxonomic Classification of DNA Sequences Using Artificial Neural Networks

### 2025-03-17 article/main.tex
article: add abstract, comment unnecessary sections, fix formula, fix typos, fix tables, rerender images with correct abbreviations

\usepackage{xcolor}
%\usepackage{tcolorbox}     % Color boxes
\usepackage{amsmath}       % Basic mathematical typesetting
\usepackage{amssymb}       % Advanced math symbols
\usepackage{amsthm}        % Theorems typesetting
%\usepackage{array}         % Advanced table column formats
%\usepackage{enumitem}      % Itemize/enumrate
%\usepackage{fancyhdr}      % Custom header/footer styles
%\usepackage{fourier}       % Adobe Utopia font
\usepackage{graphicx}      % Enhanced images support
%\usepackage{ifluatex}      % LuaTeX-specific options
%\usepackage{kantlipsum}    % English kantian-style lipsum
%\usepackage{lipsum}        % Lorem ipsum
\usepackage{listings}      % Code listings
%\usepackage{longtable}     % Multi-page tables
%\usepackage{multirow}      % Advanced table cells
%\usepackage{setspace}      % Set space between lines
%\usepackage{scrextend}     % Allows \addmargin environment
%\usepackage{tablefootnote} % Table footnotes
%\usepackage{tocloft}       % Custom ToC/LoF/LoT
%\usepackage{url}           % URL-sensitive line breaks
%\usepackage{xspace}        % Remove duplicated spaces
\usepackage{tabularx}

@@

    \abstract{Up to 350 words: background, results, conclusions}

@@

        \subsection{Motivation}
            The time required for taxonomic classification of large sequence datasets using traditional tools, such as BLAST, can span several hours, even for relatively small collections of sequences e.g. 4096 sequences. In many fields, artificial neural networks (ANN) have demonstrated the ability to outperform classical heuristic algorithms, offering faster and more accurate results. By leveraging the capabilities of ANN, this works aims to accelerate taxonomic clasisfication process while maintaining at least the same level of quality as traditional methods.

@@

        \subsection{Main Contributions}
            This paper introduces a novel artificial neural network model for calculating dissimilarities between genetic sequences, which can be used in clustering algorithm to select representative sequences for taxonomic classification. We provide all the necessary components for training the model, including dataset preparation, batch process, a loss function, and a training loop. The paper also includes an application that can perform taxonomic classification with specified parameters. Furthermore, we develope a library that allows users to compose the taxonomic classification pipelines.

        \subsection{Structure of the Paper}
            The rest of the paper is organized as follows. Section 2 provides a description of related work, reviewing previous research on taxonomic classification and highlights a research gap. Section 3 defines the problem, presents the proposed approach, and provides detailed implementation details. Section 4 describes the experimental setup, including used datasets, the performance metrics for evaluation, and the results obtained from experiments with analysis. Section 5 interprets the results, discusses the limtiations of proposed approach, and explores possible improvements for future work. Finally, Section 6 concludes the paper by summarizing the main findings and suggesting potential directions for future research.

@@

                where:
                \begin{align*}
                    A, B -& \text{compared sequences}, \\
                    n, m -& \text{lengths of sequences } A \text{ and } B, \\
                    D -& \text{similarity matrix of size } n + 1 \text{ x } m + 1, \\
                    g \in \mathbb{R} -& \text{penalty for a gap}, \\
                    s(A_i, B_j) \in \mathbb{R} -& \text{similarity between the } i\text{-th element in sequence A and the } j\text{-th element in sequence B}.
                \end{align*}

@@

@@

@@

            To compare the quality of multiple taxonomic clasisfication runs, the weighted average quality, defined by equation~\ref{Equation:WeightedAverageQuality}, was used.

@@

                \begin{table}\centering
                    \caption{Taxonomic classification execution time.}\label{Table:Experiment:Duration}
                    \begin{tabularx}{\textwidth}{|c||R|R|R|R|}
                        \hline
                        \multirow{2}{*}{\textbf{Number of sequences}} & \multicolumn{4}{|c|}{\textbf{Execution time [s]}} \\ \cline{2-5}
                                        & \textbf{NP} & \textbf{NW} & \textbf{$k$-mer} & \textbf{ANN} \\ \hline \hline
                                        1 & 7107 & \textbf{6025} & 6087 & 6077\\ \hline
                                        2 & 6129 & 5994 & \textbf{5988} & 6035\\ \hline
                                        4 & 6108 & \textbf{5983} & 6024 & 6019\\ \hline
                                        8 & 6196 & 5988 & \textbf{5925} & 6014\\ \hline
                                        16 & 6290 & \textbf{5941} & 5946 & 6118\\ \hline
                                        32 & 6279 & 6035 & \textbf{5962} & 6092\\ \hline
                                        64 & 6316 & 6113 & \textbf{6107} & 6139\\ \hline
                                        128 & 6560 & 6238 & \textbf{6206} & 6233\\ \hline
                                        256 & 6753 & 6446 & \textbf{6295} & 6363\\ \hline
                                        512 & 7091 & 6972 & \textbf{6326} & 6347\\ \hline
                                        1024 & 8059 & 8743 & \textbf{6248} & 6269\\ \hline
                                        2048 & 8517 & 16461 & 6472 & \textbf{6332}\\ \hline
                                        4096 & 9433 & \textbf{-1} & 7003 & 6550\\ \hline

                    \end{tabularx}
                \end{table}

@@

                    \begin{tabularx}{\textwidth}{|c|R|R|R|}
                        \hline
                        \multirow{2}{*}{\textbf{Number of sequences}} & \multicolumn{3}{|c|}{\textbf{Method}} \\ \cline{2-4}
                        & \textbf{NW} & \textbf{$k$-mer} & \textbf{ANN} \\ \hline \hline
                        1 & \textbf{1.0} & \textbf{1.0} & \textbf{1.0}\\ \hline
                        2 & 0.27 & \textbf{0.73} & \textbf{0.73}\\ \hline
                        4 & \textbf{0.8} & 0.2 & 0.2\\ \hline
                        8 & \textbf{0.94} & 0.88 & 0.12\\ \hline
                        16 & 0.82 & 0.82 & \textbf{0.94}\\ \hline
                        32 & 0.88 & \textbf{0.92} & 0.87\\ \hline
                        64 & 0.93 & \textbf{0.94} & 0.93\\ \hline
                        128 & 0.89 & 0.87 & \textbf{0.91}\\ \hline
                        256 & 0.89 & 0.89 & \textbf{0.93}\\ \hline
                        512 & 0.73 & 0.76 & \textbf{0.94}\\ \hline
                        1024 & \textbf{0.95} & 0.93 & \textbf{0.95}\\ \hline
                        2048 & 0.92 & \textbf{0.95} & 0.92\\ \hline
                        4096 & 0.0 & 0.93 & \textbf{0.96}\\ \hline
                    \end{tabularx}
                \end{table}

@@

        \subsection{Limitations}
            \subsubsection{Experiments Duration}
                The full process of taxonomic classification of DNA sequences for each method and each experimental subset took approximately 5 days, which led to the limitation of the number of number of experiments to single run and subset size to $4096$ sequences.

>>>>

\usepackage{xcolor}
%\usepackage{tcolorbox}     % Color boxes
\usepackage{amsmath}       % Basic mathematical typesetting
\usepackage{amssymb}       % Advanced math symbols
\usepackage{amsthm}        % Theorems typesetting
%\usepackage{array}         % Advanced table column formats
%\usepackage{enumitem}      % Itemize/enumrate
%\usepackage{fancyhdr}      % Custom header/footer styles
%\usepackage{fourier}       % Adobe Utopia font
\usepackage{graphicx}      % Enhanced images support
%\usepackage{ifluatex}      % LuaTeX-specific options
%\usepackage{kantlipsum}    % English kantian-style lipsum
%\usepackage{lipsum}        % Lorem ipsum
\usepackage{listings}      % Code listings
%\usepackage{longtable}     % Multi-page tables
\usepackage{multirow}      % Advanced table cells
%\usepackage{setspace}      % Set space between lines
%\usepackage{scrextend}     % Allows \addmargin environment
%\usepackage{tablefootnote} % Table footnotes
%\usepackage{tocloft}       % Custom ToC/LoF/LoT
%\usepackage{url}           % URL-sensitive line breaks
%\usepackage{xspace}        % Remove duplicated spaces
%\usepackage{tabularx}

@@

    \abstract{The rapid increase in the availability of DNA sequence data necessitates efficient and scalable methods for taxonomic classification. Traditional tools like BLAST, although accurate, often require significant computational resources and time, limiting their applicability to large datasets. Recent advances in artificial neural networks (ANNs) demonstrate their potential to address this issue by offering faster and more accurate solutions.

    This study introduces a novel ANN-based approach for the efficient taxonomic classification of DNA sequences. The proposed method reduces the complexity of large datasets by leveraging contrastive learning within an ANN to select representative sequences. This approach enhances computational efficiency by focusing classification efforts on these representatives rather than on the entire dataset. Comparative experiments were conducted on a benchmark dataset, evaluating the proposed ANN method alongside classical techniques such as k-mer embeddings and the Needleman-Wunsch algorithm.

    Experimental results revealed that the ANN-based approach achieves the highest weighted average classification quality among all tested methods, outperforming both classical approaches. Additionally, the ANN method demonstrates competitive execution times, benefiting from parallelized computations enabled by GPU processing.

    In conclusion, the study highlights the potential of ANN-based methods to bridge the gap between precision and scalability in genetic taxonomy. Future research directions include enhancing the network’s flexibility for varying sequence lengths and exploring its application for dimensionality reduction in broader bioinformatics tasks.}

@@

        \subsection{Motivation}
            The time required for taxonomic classification of large sequence datasets using traditional tools, such as BLAST, can span several hours, even for relatively small collections of sequences e.g. 4096 sequences. In many fields, artificial neural networks (ANN) have demonstrated the ability to outperform classical heuristic algorithms, offering faster and more accurate results. By leveraging the capabilities of ANN, this works aims to accelerate taxonomic classification process while maintaining at least the same level of quality as traditional methods.

@@

        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\
        %   BEGIN REMOVED
        %       REASON: UNNECESSARY
        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\

        % \subsection{Main Contributions}
        %     This paper introduces a novel artificial neural network model for calculating dissimilarities between genetic sequences, which can be used in clustering algorithm to select representative sequences for taxonomic classification. We provide all the necessary components for training the model, including dataset preparation, batch process, a loss function, and a training loop. The paper also includes an application that can perform taxonomic classification with specified parameters. Furthermore, we developed a library that allows users to compose the taxonomic classification pipelines.

        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\
        %   END REMOVED
        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\

        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\
        %   BEGIN REMOVED
        %       REASON: UNNECESSARY
        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\

        % \subsection{Structure of the Paper}
        %     The rest of the paper is organized as follows. Section 2 provides a description of related work, reviewing previous research on taxonomic classification and highlights a research gap. Section 3 defines the problem, presents the proposed approach, and provides detailed implementation details. Section 4 describes the experimental setup, including used datasets, the performance metrics for evaluation, and the results obtained from experiments with analysis. Section 5 interprets the results, discusses the limtiations of proposed approach, and explores possible improvements for future work. Finally, Section 6 concludes the paper by summarizing the main findings and suggesting potential directions for future research.

        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\
        %   END REMOVED
        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\

@@

                where:
                \begin{align*}
                    A, B -& \text{compared sequences}, \\
                    n, m -& \text{lengths of sequences } A \text{ and } B, \\
                    D -& \text{similarity matrix of size } n + 1 \text{ x } m + 1, \\
                    g \in \mathbb{R} -& \text{penalty for a gap}, \\
                    s(A_i, B_j) \in \mathbb{R} -& \text{similarity between the } i\text{-th element in sequence A} \\ 
                                                & \text{ and the } j\text{-th element in sequence B}.
                \end{align*}

@@

        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\
        %   BEGIN REMOVED
        %       REASON: UNNECESSARY
        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\

        % \subsection{Implementation Details}
        %     The solution consists of two components. The first is a library that contains components enabling the creation of pipelines for genetic material analysis in the form of taxonomic classification. The second component is a console application, which uses the library's components to allow for the creation and configuration of processing pipelines. Additionally, a web application was created, which allows users to submit analysis tasks through a web browser.

        %     The console application was implemented using the \textit{clap} library\cite{Rust:clap}. In library the $k$-medoid algorithm was provided by the \textit{kmedoids} library\cite{Rust:kmedoids}, and the neural network was built using the \textit{burn} library\cite{Rust:burn} with the \textit{wgpu} execution environment. Web application was implemented using the \textit{axum} library\cite{Rust:axum} with \textit{tokio}\cite{Rust:axum} async environment.

        %     Additionally, the following tools were used in the work:
        %     \begin{itemize}
        %         \item \textit{cargo} as the package manager and build system in Rust;
        %         \item \textit{rustup} for automatic management of Rust versions;
        %         \item \textit{clippy} for static code analysis in Rust;
        %         \item \textit{rustfmt} for automatic formatting of Rust source code;
        %         \item \textit{cargo test} for conducting unit tests;
        %         \item \textit{git} as the version control system, enabling change tracking and code history management.
        %     \end{itemize}

        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\
        %   END REMOVED
        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\

@@

        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\
        %   BEGIN REMOVED
        %       REASON: UNNECESSARY
        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\

        % \subsection{Experimental Setup}
        %     Experiments were conducted on the KVM based virtual machine with the specification given below:

        %     \begin{itemize}
        %         \item{
        %             \textbf{Operating System:} Ubuntu 22.04 LTS;
        %         }
        %         \item{
        %             \textbf{Processor:} 4 virtual cores of Intel Core i7-6850K;
        %         }
        %         \item{
        %             \textbf{RAM Memory:} 40 GB;
        %         }
        %         \item{
        %             \textbf{Graphics Card:} Nvidia GeForce GTX 1080 TI;
        %         }
        %         \item{
        %             \textbf{Disk:} 1 TB network drive with a read speed of 1 Gbps;
        %         }
        %         \item{
        %             \textbf{Software:} \texttt{BLAST} package version 2.16.0 and graphics card drivers.
        %         }
        %     \end{itemize}

        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\
        %   END REMOVED
        % /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\

@@

            To compare the quality of multiple taxonomic classification runs, the weighted average quality, defined by equation~\ref{Equation:WeightedAverageQuality}, was used.

@@

                \begin{table}\centering
                    \caption{Taxonomic classification execution time.}\label{Table:Experiment:Duration}
                    \begin{tabular}{|c||c|c|c|c|}
                        \hline
                        \multirow{2}{*}{\textbf{Number of sequences}} & \multicolumn{4}{|c|}{\textbf{Execution time [s]}} \\
                                        & \textbf{NP} & \textbf{NW} & \textbf{$k$-mer} & \textbf{ANN} \\ \hline \hline
                                        1 & 7107 & \textbf{6025} & 6087 & 6077\\ \hline
                                        2 & 6129 & 5994 & \textbf{5988} & 6035\\ \hline
                                        4 & 6108 & \textbf{5983} & 6024 & 6019\\ \hline
                                        8 & 6196 & 5988 & \textbf{5925} & 6014\\ \hline
                                        16 & 6290 & \textbf{5941} & 5946 & 6118\\ \hline
                                        32 & 6279 & 6035 & \textbf{5962} & 6092\\ \hline
                                        64 & 6316 & 6113 & \textbf{6107} & 6139\\ \hline
                                        128 & 6560 & 6238 & \textbf{6206} & 6233\\ \hline
                                        256 & 6753 & 6446 & \textbf{6295} & 6363\\ \hline
                                        512 & 7091 & 6972 & \textbf{6326} & 6347\\ \hline
                                        1024 & 8059 & 8743 & \textbf{6248} & 6269\\ \hline
                                        2048 & 8517 & 16461 & 6472 & \textbf{6332}\\ \hline
                                        4096 & 9433 & \textbf{-1} & 7003 & 6550\\ \hline

                    \end{tabular}
                \end{table}

@@

                    \begin{tabular}{|c|c|c|c|}
                        \hline
                        \multirow{2}{*}{\textbf{Number of sequences}} & \multicolumn{3}{|c|}{\textbf{Method}} \\
                        & \textbf{NW} & \textbf{$k$-mer} & \textbf{ANN} \\ \hline \hline
                        1 & \textbf{1.0} & \textbf{1.0} & \textbf{1.0}\\ \hline
                        2 & 0.27 & \textbf{0.73} & \textbf{0.73}\\ \hline
                        4 & \textbf{0.8} & 0.2 & 0.2\\ \hline
                        8 & \textbf{0.94} & 0.88 & 0.12\\ \hline
                        16 & 0.82 & 0.82 & \textbf{0.94}\\ \hline
                        32 & 0.88 & \textbf{0.92} & 0.87\\ \hline
                        64 & 0.93 & \textbf{0.94} & 0.93\\ \hline
                        128 & 0.89 & 0.87 & \textbf{0.91}\\ \hline
                        256 & 0.89 & 0.89 & \textbf{0.93}\\ \hline
                        512 & 0.73 & 0.76 & \textbf{0.94}\\ \hline
                        1024 & \textbf{0.95} & 0.93 & \textbf{0.95}\\ \hline
                        2048 & 0.92 & \textbf{0.95} & 0.92\\ \hline
                        4096 & 0.0 & 0.93 & \textbf{0.96}\\ \hline
                    \end{tabular}
                \end{table}

@@

        \subsection{Limitations}
            \subsubsection{Experiments Duration}
                The full process of taxonomic classification of DNA sequences for each method and each experimental subset took approximately 5 days, which led to the limitation of the number of experiments to single run and subset size to $4096$ sequences.
